Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-05-19 doc/manuscript/translocation_bioarxiv_SI.tex
Address formatting comments on compiled manuscript

In the Results section of the main paper, we describe the stringent set
of outlier SNPS (identified using a Bonferroni-corrected p-value of
0.01). The liberal set, identified using a Bonferroni-corrected p-value
of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct
genes across 16 contigs. Our analysis found no significantly
over-represented GO terms in this set of 30 genes, or in the set of 38
outlier regions identified with the splined window analysis.

In the comparison of individual outlier variants between naive and
recovering populations, we detected consistent allele frequency changes
across all genetic clusters for 1 of 7 of the associated genes (RIN3;
Fig. 1). However, for the other 6 genes we
still found a statistically significant signal of selection in 2 of the
3 genetic clusters (containing populations 6--9;
Fig. 1, Figure~\ref{fig-selectionresults}
A). Therefore, although some outlier variant associations have a more
limited geographic extent than RIN3, they still describe parallel
evolutionary changes following Bd exposure and are therefore interesting
candidates for future studies.

@@

Swab extracts were analyzed singly instead of in triplicate
\citep{kriger2006}. For analysis of swabs collected during 2005--2014,
we used standards developed from known concentrations of zoospores
\citep{boyle2004}, and after 2014, we used standards based on single
ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
between samples analyzed using both types of standards, Bd in the study
area has an average of 60 ITS1 copies per zoospore. To express all qPCR
results as the number of ITS1 copies, starting quantities obtained using
the zoospore standard (measured as ``zoospore equivalents'') were
multiplied by 60. In addition, all qPCR quantities (regardless of
standard) were multiplied by 80 to account for the fact that DNA
extracts from swabs were diluted 80-fold during extraction and PCR
\citep{vredenburg2010}.

@@

We extracted yearly survival probabilities for translocated adults
\(\sigma_{A_T}\) and naturally recruited adults \(\sigma_{A_R}\) from
the CMR model used to estimate population-level parameters of
translocated populations (see ``Estimation of adult survival and
population size'' above). Although we observed yearly variability in
adult survival within a population, the magnitude of this variability
was small compared to among population variability
(Fig. 3). Thus, we did not include
yearly within-population variability in adult survival in this analysis.
However, within a population there was substantial yearly variability in
the successful recruitment of adults, greater than what we would expect
from Poisson variability around a mean value. Therefore, we allowed for
yearly variability in the probability of successful recruitment
\(\omega\) (additional details provided in ``Estimating model
parameters'' below).

@@

The baseline parameter values for the model and how they were estimated
are given in Table~\ref{tbl-param_values}. Parameters \(\sigma_{A_T}\)
and \(\sigma_{A_R}\) were extracted directly from our CMR model
described previously in this study. For populations where we had a
sufficient number of PIT-tagged, naturally-recruited adults, we observed
that \(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be notably different,
with \(\sigma_{A_R} > \sigma_{A_T}\)
(Figure~\ref{fig-compare_surv_probs}). For populations lacking
sufficient numbers of naturally-recruited adults, we were unable to
directly estimate \(\sigma_{A_R}\), and instead set
\(\sigma_{A_R} = \sigma_{A_T}\).

@@

\caption{\label{fig-selectionresults}Evidence for selection on
individual variants in recovering MYL frog populations. (A) PCA
calculated from binary SNPs showing the genomic relationship of samples.
Numeric labels and colors match those in
Fig. 1 A. Populations 1-7 are \emph{R.
sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B) qqplot
showing observed and expected p-values for 148,307 SNPs and INDELS.
Dashed line shows p-value that identifies outliers. (C) Manhattan plot
showing p-value for each SNP. SNPs are sorted by genomic position and
contigs are sorted by size. Red line shows p-value that identifies
outliers. Outlier SNPs above this threshold are highlighted and labeled.
Bold labels indicate the presence of at least one non-synonymous SNP in
that gene.}

@@

\caption{\label{fig-bdload-beforeafter}Evidence that frogs translocated
from recovering populations can maintain the benefits of
resistance/tolerance in non-natal habitats. For frogs translocated to
each of the 12 recipient sites, Bd loads for the period immediately
prior to translocation versus during the 1-year period after
translocation. Bd loads are expressed as the number of ITS1 copies per
skin swab, as estimated by qPCR of the Bd ITS1 region. Loads indicative
of severe disease are \textgreater{} 5.8 ITS copies (on a
log\textsubscript{10} scale; see SI Appendix text for details). Box plot
details are as in Fig. 4.}

@@

\caption{\label{fig-survival-postdens}Evidence that Bd load is not an
important predictor of post-translocation frog survival. For the best
model identifying predictors of 1-year frog survival following
translocation, estimated posterior density curves and shaded 95\%
uncertainty intervals for the intercept and all predictor variables. In
the Bayesian framework in which the model was developed, variables are
considered important predictors if the associated uncertainty interval
does not overlap zero (indicated by the dashed red line).}

@@

\caption{\label{fig-compare_surv_probs}Evidence that frog
resistance/tolerance in non-natal habitats is maintained across
generations. Comparison of the average yearly adult survival
probabilities for adults translocated to each of 3 sites versus adults
that were naturally recruited at each site (as a result of reproduction
by translocated frogs). In contrast to
Fig. 3, these are not survival
probabilities from the first year following translocation, but instead
represent averaged survival probabilities across multiple years and
cohorts. Points are median estimates and error bars give the 95\%
uncertainty intervals around the estimates, accounting for yearly
variation in survival probabilities. The large error bars for 70550
reflect the large within and among individual variability in yearly
survival probability at this site (as also reflected in one-year
survival probabilities in Fig. 3). All
estimates are derived from the mrmr package.}

>>>>

In the Results section of the main paper, we describe the stringent set
of outlier SNPs (identified using a Bonferroni-corrected p-value of
0.01). The liberal set, identified using a Bonferroni-corrected p-value
of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct
genes across 16 contigs. Our analysis found no significantly
over-represented GO terms in this set of 30 genes, or in the set of 38
outlier regions identified with the splined window analysis.

In the comparison of individual outlier variants between naive and
recovering populations, we detected consistent allele frequency changes
across all genetic clusters for only 1 of 7 of the associated genes
(RIN3; Fig. 1). However, for the other 6
genes we still found a statistically significant signal of selection in
2 of the 3 genetic clusters (containing populations 6--9;
Fig. 1, Figure~\ref{fig-selectionresults}
A). Therefore, although some outlier variant associations have a more
limited geographic extent than RIN3, they still describe parallel
evolutionary changes following Bd exposure and are therefore interesting
candidates for future studies.

@@

Swab extracts were analyzed using standard Bd DNA extraction and qPCR
methods \citep{boyle2004}, and extracts were analyzed singly instead of
in triplicate \citep{kriger2006}. For analysis of swabs collected during
2005--2014, we used standards developed from known concentrations of
zoospores \citep{boyle2004}, and after 2014, we used standards based on
single ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
between samples analyzed using both types of standards, Bd in the study
area has an average of 60 ITS1 copies per zoospore. To express all qPCR
results as the number of ITS1 copies, starting quantities obtained using
the zoospore standard (measured as ``zoospore equivalents'') were
multiplied by 60. In addition, all qPCR quantities (regardless of
standard) were multiplied by 80 to account for the fact that DNA
extracts from swabs were diluted 80-fold during extraction and PCR
\citep{vredenburg2010}.

@@

We extracted yearly survival probabilities for translocated adults
\(\sigma_{A_T}\) and naturally recruited adults \(\sigma_{A_R}\) from
the CMR model used to estimate population-level parameters of
translocated populations (see ``Materials and Methods - Estimation of
frog survival and abundance'' in main paper). Although we observed
yearly variability in adult survival within a population, the magnitude
of this variability was small compared to among-population variability
(Fig. 3). Thus, we did not include
yearly within-population variability in adult survival in this analysis.
However, within a population there was substantial yearly variability in
the successful recruitment of adults, greater than what we would expect
from Poisson variability around a mean value. Therefore, we allowed for
yearly variability in the probability of successful recruitment
\(\omega\) (additional details provided in ``Estimating model
parameters'' below).

@@

The baseline parameter values for the model and how they were estimated
are given in Table~\ref{tbl-param_values}. Parameters \(\sigma_{A_T}\)
and \(\sigma_{A_R}\) were extracted directly from our CMR models (see
``Materials and Methods - Estimation of frog survival and abundance'' in
main paper). For populations where we had a sufficient number of
PIT-tagged, naturally-recruited adults, we observed that
\(\sigma_{A_T}\) and \(\sigma_{A_R}\) could be notably different, with
\(\sigma_{A_R} > \sigma_{A_T}\) (Figure~\ref{fig-compare_surv_probs}).
For populations lacking sufficient numbers of naturally-recruited
adults, we were unable to directly estimate \(\sigma_{A_R}\), and
instead set \(\sigma_{A_R} = \sigma_{A_T}\).

@@

\caption{\label{fig-selectionresults}Evidence for selection on
individual variants in recovering MYL frog populations. (A) PCA
calculated from binary SNPs showing the genomic relationship of samples.
Numeric labels and colors match those in
Fig. 1A. Populations 1-7 are \emph{R.
sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B) qqplot
showing observed and expected p-values for 148,307 SNPs and INDELS.
Dashed line shows p-value that identifies outliers. (C) Manhattan plot
showing p-value for each SNP. SNPs are sorted by genomic position and
contigs are sorted by size. Red line shows p-value that identifies
outliers. Outlier SNPs above this threshold are highlighted and labeled.
Bold labels indicate the presence of at least one non-synonymous SNP in
that gene.}

@@

\caption{\label{fig-bdload-beforeafter}Evidence that frogs translocated
from recovering populations can maintain the benefits of
resistance/tolerance in non-natal habitats. For frogs translocated to
each of the 12 recipient sites, Bd loads for the period immediately
prior to translocation versus during the 1-year period after
translocation. Bd loads are expressed as the number of ITS1 copies per
skin swab, as estimated by qPCR of the Bd ITS1 region. Box plots show
medians, first and third quartiles, largest and smallest values within
1.5x interquartile range, and values outside the 1.5x interquartile
range. Loads indicative of severe disease are \textgreater{} 5.8 ITS
copies (on a log\textsubscript{10} scale).}

@@

\caption{\label{fig-survival-postdens}Evidence that Bd load is not an
important predictor of post-translocation frog survival. For the best
model identifying predictors of 1-year frog survival following
translocation, estimated posterior density curves and shaded 95\%
uncertainty intervals for the intercept and all predictor variables. In
the Bayesian framework in which the model was developed, variables are
considered important predictors if the associated uncertainty interval
does not overlap zero (indicated by the dashed red line). Predictor
variables shown on the y-axis are defined as follows: length = frog
size, snow\_t = winter severity in the year of translocation (measured
on April 1), snow\_t1 = winter severity in the year following
translocation (measured on April 1), day = day of year on which a
translocation was conducted, bd\_load = Bd load, elevation = site
elevation, order\_first = within-site translocation order, sex\_male =
frog sex, donor\_70567 and donor\_72996 = donor population.}

@@

\caption{\label{fig-compare_surv_probs}Evidence that frog
resistance/tolerance in non-natal habitats is maintained across
generations. Comparison of the average yearly adult survival
probabilities for adults translocated to each of 3 sites versus adults
that were naturally recruited at each site (as a result of reproduction
by translocated frogs). In contrast to
Fig. 3, these are not survival
probabilities from the first year following translocation, but instead
represent averaged survival probabilities across multiple years and
cohorts. Points are median estimates and error bars give the 95\%
uncertainty intervals around the estimates, accounting for yearly
variation in survival probabilities. All estimates were derived using
the mrmr package.}

### 2023-05-19 doc/manuscript/translocation_pnas.tex
Address formatting comments on compiled manuscript

Using the liberal set of variant outliers, we performed a statistical
overrepresentation test (Fisher's exact test) using Blast2GO
\citep{conesa2005}. First, we retrieved the BLAST hits and mapped GO
terms for each gene in our targeted transcriptome. Next, we compared the
30 unique outlier genes from the liberal set of outlier variants to the
complete set of genes in our target transcriptome to determine if any GO
biological functions, molecular functions, or cellular processes were
overrepresented. We then repeated this process for the set of 33 genes
located in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
splined windows. Code used for genomic analyses and to create figures is
available from the following GitHub repository:
\url{https://github.com/allie128/mylf-selection}.

@@

We conducted 1--4 translocations per site (Figure S3) and
each translocated cohort included 18 to 99 frogs (median = 30). In
preparation for each translocation, adult frogs were collected from the
donor population and measured, weighed, swabbed, and PIT tagged. Frogs
were transported to the recipient site either on foot or via helicopter.
Following release, translocated populations were visited approximately
once per month during the summer active season (median number of visits
per summer = 2; range = 1--10) and assessed using diurnal CMR surveys
and VES. CMR surveys allowed estimation of adult survival, adult
population size, and recruitment of new adults, and VES provided
estimates of tadpole and juvenile abundance. During 2006-2012, CMR
surveys were conducted on a single day (primary period) per site visit,
during which all habitats were searched repeatedly for adult frogs.
Frogs were captured using handheld nets, identified via their PIT tag
(or tagged if they were untagged), measured, weighed, swabbed, and
released at the capture location. During 2013-2022, we generally used a
robust design in which all habitats were searched during several
consecutive days (median number of secondary periods per primary period
= 3; range = 3--7), and frogs were processed as described above.
However, when the number of frogs detected on the first survey day was
zero or near zero, we typically conducted only a single-day CMR survey.
When using a robust design, within a primary period, frogs that were
captured during more than one secondary period were measured, weighed,
and swabbed during the first capture, and during subsequent captures
were only identified and released. Frog skin swabs were analyzed using
standard Bd DNA extraction and qPCR methods. For details of survey and
swab analysis methods, see \citep{joseph2018}.

@@

For each of the translocated populations, we used the presence of
tadpoles and/or juveniles from VES and captures of new recruits (i.e,
untagged adults) in CMR surveys to provide two measures of successful
reproduction. To calculate the proportion of years in which
tadpoles/juveniles were present at a site, we excluded surveys conducted
in the year of the initial translocation to that site. This exclusion
accounted for the fact that all translocations were conducted after the
breeding period and reproduction would therefore not have occurred until
the following year. Similarly, to calculcate the proportion of years in
which new recruits were present at a site, we excluded surveys conducted
during the 3 years following the initial translocation. This accounted
for the multi-year tadpole and juvenile stages in MYL frogs
(Table S2).

@@

We conducted two analyses. In the first, we addressed the more general
question of whether survival of frogs in the first translocation to a
site predicted the survival of frogs in subsequent translocations.
Group-level effects included site\_id or translocation\_id. The second
analysis focused specifically on identifying predictors of
post-translocation frog survival using a wide range of variables
describing characteristics of frogs, sites, and translocations
(predictor variables: Bd load (bd\_load), sex (sex\_male), frog size
(length), site elevation (elevation), winter severity in the year of
translocation (snow\_t), winter severity in the year following
translocation (snow\_t1), donor population (donor\_72996), day of year
on which a translocation was conducted (day), and translocation order
(order\_first)). In both analyses we used 1-year survival as a single
metric of survival that was available for all translocated populations
regardless of how many years of post-translocation data were available
(range = 3--16). Group-level effects included site\_id,
translocation\_id, or translocation\_id nested within site\_id. In both
analyses, the response variable, probability of 1-year frog survival,
was rounded to integer values to produce a binomial outcome, and all
models used a bernoulli family. For all models, we used default priors,
four chains, and 5000 iterations each for warmup and post-warmup. We
checked convergence of the MCMC algorithm using trace plots and Rhat.

@@

After parameterizing our model with CMR-estimated adult frog survival
probabilities and other known vital rates
(Table S2), we performed four analyses (described in
detail in the SI Appendix). First, we compute the long-run growth rate
\(\lambda\) for each of our 12 translocated populations to determine if
the populations were deterministically predicted to grow or decline in
the long-run. Second, we compute the elasticity of \(\lambda\) to model
parameters, to quantify how much changes in adult survival, for example,
affected the long-run growth rate. Third, we included demographic and
environmental stochasticity in our model and simulated the 50-year
viability (i.e., 1 - extinction probability) of populations given an
introduction of 40 adult individuals into an unoccupied habitat.
Finally, we fit our model to our longest translocation trajectory to
confirm that our model could reasonably reproduce the observed recovery
trajectories of MYL frogs following reintroductions. Additional details
are provided in the SI Appendix. Code to replicate the analyses can be
found at \url{https://github.com/SNARL1/translocation}.}

@@

@@

\caption{\label{fig-viability}Population viability analysis of
translocated MYL frogs quantified the long-run growth rate and
extinction probability of frog populations: (A) Predicted long-run
growth rate \(\lambda\) for different values of yearly adult survival
probability \(\sigma_{A_R}\) and probability of recruitment \(\omega\),
given the parameterized, deterministic model. Colored points show the
predicted \(\lambda\) values for the twelve translocated populations
when probability of recruitment into the adult class is \(\omega = 0.3\)
(indicated by the dashed line). The red line shows where
\(\lambda = 1\). (B) Predicted 50-year extinction probabilities of the
12 translocated populations, given demographic stochasticity,
environmental variability in \(\omega\), and different mean values of
\(\omega\). The dashed line shows where \(\omega = 0.3\), for comparison
to (A). There are 6 lines at extinction probability = 1, 5 of which
(70414--70619) are hidden beneath the line for 70505. (C) 100 simulated
trajectories (black lines) from the population viability model that most
closely matched the observed abundance trajectory of adult amphibians at
site 70550 (light orange).}

>>>>

Using the liberal set of outlier variants, we performed a statistical
overrepresentation test (Fisher's exact test) using Blast2GO
\citep{conesa2005}. First, we retrieved the BLAST hits and mapped GO
terms for each gene in our targeted transcriptome. Next, we compared the
30 unique outlier genes from the liberal set of outlier variants to the
complete set of genes in our target transcriptome to determine if any GO
biological functions, molecular functions, or cellular processes were
overrepresented. We then repeated this process for the set of 33 genes
located in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
splined windows. Code used for genomic analyses and to create figures is
available from the following GitHub repository:
\url{https://github.com/allie128/mylf-selection}.

@@

We conducted 1--4 translocations per site (Figure S3,
Figure~\ref{fig-translocation-survival}) and each translocated cohort
included 18 to 99 frogs (median = 30). In preparation for each
translocation, adult frogs were collected from the donor population and
measured, weighed, swabbed, and PIT tagged. Frogs were transported to
the recipient site either on foot or via helicopter. Following release,
translocated populations were visited approximately once per month
during the summer active season (median number of visits per summer = 2;
range = 1--10) and assessed using diurnal CMR surveys and VES. CMR
surveys allowed estimation of adult survival, adult population size, and
recruitment of new adults, and VES provided estimates of tadpole and
juvenile abundance. During 2006-2012, CMR surveys were conducted on a
single day (primary period) per site visit, during which all habitats
were searched repeatedly for adult frogs. Frogs were captured using
handheld nets, identified via their PIT tag (or tagged if they were
untagged), measured, weighed, swabbed, and released at the capture
location. During 2013-2022, we generally used a robust design in which
all habitats were searched during several consecutive days (median
number of secondary periods per primary period = 3; range = 3--7), and
frogs were processed as described above. However, when the number of
frogs detected on the first survey day was zero or near zero, we
typically conducted only a single-day CMR survey. When using a robust
design, within a primary period, frogs that were captured during more
than one secondary period were measured, weighed, and swabbed during the
first capture, and during subsequent captures were only identified and
released.

@@

For each of the translocated populations, we used the presence of
tadpoles and/or juveniles from VES and counts of new recruits (i.e,
untagged adults) in CMR surveys to provide two measures of successful
reproduction. To calculate the proportion of years in which
tadpoles/juveniles were present at a site, we excluded surveys conducted
in the year of the initial translocation to that site. This exclusion
accounted for the fact that all translocations were conducted after the
breeding period and reproduction would therefore not have occurred until
the following year. Similarly, to calculcate the proportion of years in
which new recruits were present at a site, we excluded surveys conducted
during the 3 years following the initial translocation. This accounted
for the multi-year tadpole and juvenile stages in MYL frogs
(Table S2).

@@

We conducted two analyses. In the first, we addressed the more general
question of whether survival of frogs in the first translocation to a
site predicted the survival of frogs in subsequent translocations.
Group-level effects included site\_id or translocation\_id. The second
analysis focused specifically on identifying predictors of
post-translocation frog survival using a wide range of variables
describing characteristics of frogs, sites, and translocations
(predictor variables: Bd load, sex, frog size, site elevation, winter
severity in the year of translocation, winter severity in the year
following translocation, donor population, day of year on which a
translocation was conducted, and translocation order). In both analyses
we used 1-year survival as a single metric of survival that was
available for all translocated populations regardless of how many years
of post-translocation data were available (range = 3--16). Group-level
effects included site\_id, translocation\_id, or translocation\_id
nested within site\_id. In both analyses, the response variable,
probability of 1-year frog survival, was rounded to integer values to
produce a binomial outcome, and all models used a bernoulli family. For
all models, we used default priors, four chains, and 5000 iterations
each for warmup and post-warmup. We checked convergence of the MCMC
algorithm using trace plots and Rhat.

@@

After parameterizing our model with CMR-estimated adult frog survival
probabilities and other known vital rates
(Table S2), we performed four analyses. First, we
compute the long-run growth rate \(\lambda\) for each of our 12
translocated populations to determine if the populations were
deterministically predicted to grow or decline in the long-run. Second,
we compute the elasticity of \(\lambda\) to model parameters, to
quantify how much changes in adult survival, for example, affected the
long-run growth rate. Third, we included demographic and environmental
stochasticity in our model and simulated the 50-year viability (i.e., 1
- extinction probability) of populations given an introduction of 40
adult individuals into an unoccupied habitat. Finally, we fit our model
to our longest translocation trajectory to confirm that our model could
reasonably reproduce the observed recovery trajectories of MYL frogs
following reintroductions. Additional details are provided in the SI
Appendix. Code to replicate the analyses can be found at
\url{https://github.com/SNARL1/translocation}.

@@

\clearpage

@@

\caption{\label{fig-viability}Population viability analysis of
translocated MYL frogs quantified the long-run growth rate and
extinction probability of frog populations: (A) Predicted long-run
growth rate \(\lambda\) for different values of yearly adult survival
probability \(\sigma_{A_R}\) and probability of recruitment \(\omega\),
given the parameterized, deterministic model. Colored points show the
predicted \(\lambda\) values for the twelve translocated populations
when probability of recruitment into the adult class is \(\omega = 0.3\)
(indicated by the dashed line). The red line shows where
\(\lambda = 1\). Note that the point for 70413 is mostly hidden behind
other points. (B) Predicted 50-year extinction probabilities of the 12
translocated populations, given demographic stochasticity, environmental
variability in \(\omega\), and different mean values of \(\omega\). The
dashed line shows where \(\omega = 0.3\), for comparison to (A). There
are 6 lines at extinction probability = 1, 5 of which (70414--70619) are
hidden beneath the line for 70505. (C) 100 simulated trajectories (black
lines) from the population viability model that most closely matched the
observed abundance trajectory of adult amphibians at site 70550 (light
orange).}
